Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-03 talk/dls2014/paper/paper.tex
Why no space before *this* section title??
more notes

\subsection{Setup\label{sub:Setup}}

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in Section~\ref{sec:performance-bench}. We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT is a tracing JIT~\cite{cfbolz09}
\remi{explain?}  running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

@@

@@

@@

For PyPy-STM the average memory requirement is 21~MiB and there are
$\sim 10$ major collections during the runtime. Normal PyPy with a GIL
grows its memory up to just 7~MiB and does not do a single major
collection in that time. Compared to normal PyPy, we are missing a
memory optimisation to store small objects in a more compact
way. Additionally, since normal PyPy uses a GIL, it does not need to
duplicate any data structures like e.g.\ the Nursery for each
thread. This, the missing optimisation, and the additional memory
requirements for STM explained above account for this difference.
We expect to improve this aspect in the future; in this paper we
want to focus first on performance.
\remi{I don't know how much sense it makes to go deeper. We will
improve this in the future, but right now this is the overall picture.}

@@

@@

We use coarse-grained locking for the first three benchmarks and no
locking for the last three. This is important because Jython, which
uses fine-grained locking instead of a GIL, is only expected to scale
with the number of threads for the latter group. It is not able to
scale when using coarse-grained locking. STM, however, uses atomic
blocks instead of a single lock to synchronise accesses to the
shared data structures. Since atomic blocks map to transactions,
our STM system may still be able to get a speedup on more than one
thread by running transactions in parallel.

@@

\emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
however, it never even beats CPython, the reference implementation of
Python. This means that without the JIT, our performance is not
competitive. So we will now look at how well our system works when we
enable the JIT.

@@

\paragraph{JIT benchmarks:} The speedups from enabling the JIT in
these benchmarks range from $10-50\times$. This is why we had to do
without Jython and CPython here, since they would be much further up
in the plots. Also, in order to get more stable results, we increased
the input size of all benchmarks to get reasonable execution times.

The results are presented in Figure~\ref{fig:performance-nojit}. We
see that the performance is much less stable. There is certainly more
work required in this area. The slowdown factor for switching from GIL
to STM ranges around \remi{$1-2.4\times$}, and we beat GIL performance
in half of the benchmarks.

>>>>

\medskip   % why, Latex, why??
\subsection{Setup\label{sub:Setup}}

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in Section~\ref{sec:performance-bench}.
\cfbolz{the details about why the JIT confuses everything should go into
Section~\ref{sec:performance-bench}. here just write "because the integration
with the JIT is incomplete"}
We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT is a tracing JIT~\cite{cfbolz09}
\remi{explain?}  running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

@@

\cfbolz{can you explain how you measure the errors?}

@@

\cfbolz{part of this section is wrong here. maybe in the low-level section? here you should only measure}

@@

For PyPy-STM the average memory requirement is 21~MiB and there are
$\sim 10$ major collections during the runtime. Normal PyPy with a GIL
grows its memory up to just 7~MiB and does not do a single major
collection in that time. Compared to normal PyPy, we are missing a
memory optimisation to store small objects in a more compact
way. Additionally, since normal PyPy uses a GIL, it does not need to
duplicate any data structures like e.g.\ the Nursery for each
thread. This, the missing optimisation, and the additional memory
requirements for STM explained above account for this difference.
We expect to improve this aspect in the future; in this paper we
want to focus first on performance.
\remi{I don't know how much sense it makes to go deeper. We will
improve this in the future, but right now this is the overall picture.}
\cfbolz{you can't kill it, otherwise a reviewer will ask. also, the pypy-no-stm numbers are meaningless as long as the options are different}

@@

\cfbolz{why is the error so big with 4 threads?}

@@

We use coarse-grained locking for the first three benchmarks and no
locking for the last three. This is important because Jython, which
uses fine-grained locking instead of a GIL, is only expected to scale
with the number of threads for the latter group. It is not able to
scale when using coarse-grained locking. STM, however, uses atomic
blocks instead of a single lock to synchronise accesses to the
shared data structures. Since atomic blocks map to transactions,
our STM system may still be able to get a speedup on more than one
thread by optimistically running transactions in parallel.

@@

\emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
however, it does not consistently beat CPython, the reference implementation of
Python. This means that without the JIT, our performance is not
competitive. So we will now look at how well our system works when we
enable the JIT.

@@

\cfbolz{I suggest making the rest a new section}
\paragraph{JIT benchmarks:} The speedups from enabling the JIT in
these benchmarks range from $10-50\times$. This is why we had to do
without Jython and CPython here, since they would be much further up
in the plots. Also, in order to get more stable results, we increased
the input size of all benchmarks to get reasonable execution times.

The results are presented in Figure~\ref{fig:performance-jit}. We
see that the performance is much less stable. There is certainly more
work required in this area. The slowdown factor for switching from GIL
to STM ranges around \remi{$1-2.4\times$}, and we beat GIL performance
in half of the benchmarks.
